Fail Faster

### 2025-03-25 paper/paper.tex
A few quick fixes, plus polishing on last two sections

let int_pair_inlined : int -> SR.t -> int * int =
  fun _ sr ->
    let x = SR.int sr ~lo:0 ~hi:100 in
    let y = SR.int sr ~lo:0 ~hi:x in
    (x,y)
\end{lstlisting}
\caption{Simplifying a Generator in OCaml}\label{fig:simplify-ocaml}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
def intPair : Gen[(Long,Long)] = for {
  x <- Gen.choose(0,1000)
  y <- Gen.choose(0,x)
} yield (x,y)

def intPairInlined : (Gen.Parameters, Seed) => 
(Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
  }
\end{lstlisting}
\caption{Simplifying a Generator in Scala}\label{fig:simplify-scala}
  \end{subfigure}
  \caption{Simplifying Generators}\label{fig:simplify}
\end{figure}

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking bug-finding speed}
To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
case studies using Etna~\cite{etna}.
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test a specific aspect of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations).
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.
Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
expect the results in this section to extend to \scalaname.

@@

In the years since the original QuickCheck paper~\cite{quickcheck00}, PBT has had remarkable success
in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages---including non-strict or procedural ones---might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.

@@

  \textbf{Haskell} is home to the original PBT implementation, and
  so it is natural to ask why we have not built a \name{} in it yet. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
  complex enough programs however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system called Template Haskell~\cite{templatehaskell}, which is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to build a \name{} version of Haskell
  QuickCheck, but the benefits would be much less pronounced.

  \textbf{F\#} has both multi-stage programming capabilities~\cite{fsmeta} and a well-used PBT library called FsCheck~\cite{fscheck}.
  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building an \name{}
  in F\# is primarily an engineering effort.
  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations

  \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
  structurally encourage monadic eDSLs---it does not have a special monadic syntax---and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.

@@

Reordering test input sampling is perhaps the most well-studied way to accelerate bug-finding in PBT.
 Feedback-driven mechanisms
like Targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
Coverage-Guided PBT~\cite{lampropoulosCoverageGuidedProperty2019} have the
potential to speed up testing by changing the generation order to try to find
``interesting'' inputs faster. Separately, enumerative approaches to
PBT~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
try to get to bugs faster by leveraging the ``small-scope hypothesis'' and
attempting to quickly exhaust the list of small inputs to the program under
test. All of these approaches can have significant performance benefits in
practice, but they are largely orthogonal to our contributions; we expect
\name{} could be used to speed these existing techniques to varying degrees.

Similarly orthogonal to our approach are techniques for quickly filtering out
inputs that are invalid for testing. Some of these approaches work statically,
by automatically deriving generators that produce valid inputs by
construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
filter dynamically via laziness~\cite{claessenGeneratingConstrainedRandom2015}
or by solving satisfiability
problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
In all cases, these approaches could be further improved by insights from \name.

The most direct comparison to our work is
QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}, an
implementation of Haskell's QuickCheck that exploits the inherent parallelism of
PBT to achieve significant performance gains. QuickerCheck, like \name, was
designed with performance engineering in mind, taking seriously the idea that
PBT is a performance critical task. Still, despite their similar motivations and
considerations, the solutions are entirely different and complementary.

\paragraph{Multi-Stage Programming}
% \jwc{A generator is a parser of
% randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
% lots of staged parser libraries!}
Staging's roots come from quasiquotation in LISP \cite{bawden99}, where the
unified representation of data and code allows for sophisticated
metaprogramming. Quasiquation-style macros were introduced to the ML family of
languages by Nielson and Nielson's Two-Level Functional Languages~\cite{nielson92} and
MetaML~\cite{taha00}, which in turn quickly inspired MetaOCaml~\cite{metaocaml} (and more recently
MacoCaml~\cite{macocaml}) in OCaml, Lightweight Modular Staging in Scala~\cite{lms}, Template
Haskell in Haskell~\cite{templatehaskell}, and more.  In addition to implementations, researchers
have studied the type-theoretic foundations of multi-stage languages \cite{cmtt,
davies-pfenning, davies17}, as well as ways of combining multi-stage types with
other sophisticated features like dependent types \cite{kovacs24}. Meanwhile,
the Racket community continues the LISP tradition with a long and fruitful
line of work studying how macros and metaprogramming can be used to build extensible
DSLs \cite{flatt12, samth11, skthesis}.

Back in the ML-family world, the primary use of staging is building embedded DSLs
with minimal performance overhead, an idea which has come to be known as ``abstraction without regret''
\cite{lms,parreaux17,parreauxthesis, sheard99, trattdsl}.
This technique has been applied across a range of domains, including
Big data processing \cite{jet12}, stream processing \cite{moller20,strymonas},
query processing \cite{rhyme}, and parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap}.
This work on staged parsers is the closest analogue to ours---indeed, PBT research has drawn an explicit link between parsing and generation, framing generators
as ``parsers of randomness'' \cite{goldsteinParsingRandomness2022}.
However, this equivalence is primarily theoretical. PBT generators are (a) not actually implemented as parsers
of random sequences, and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
For this reason, we must build staged generator libraries from scratch and cannot simply use existing staged parser libraries.

@@

\section{Conclusion \& Future Work}
\label{section:concl}
In this paper, we identified, studied, and proposed potential solutions to two
important sources of inefficiency in PBT generator libraries: abstraction
overhead and choice of \rand{}.  In the future, we hope to continue to investigate and push the boundaries of
PBT generator performance.

On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion~\cite{kiselyov2022letrecinsertioneffects} that yields optimal let placement,
which we did not use in \camlname{}.
Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions. Currently, the \camlname{}
recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this
overhead can be eliminated using a type-level heterogenous list to represent the state at compile time.

On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cite{lehmerrng}, WyHash~\cite{wyrand},
and variants of xoroshiro~\cite{xorshiro}. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
or even generating random numbers ahead of time. Last, we hope to investigate the degree to which actual statistical randomness
matters in PBT.

>>>>

let int_pair_inlined : int -> SR.t -> int * int =
  fun _ sr ->
    let x = SR.int sr ~lo:0 ~hi:100 in
    let y = SR.int sr ~lo:0 ~hi:x in
    (x,y)
\end{lstlisting}
\caption{Simplifying a generator in OCaml}\label{fig:simplify-ocaml}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
def intPair : Gen[(Long,Long)] = for {
  x <- Gen.choose(0,1000)
  y <- Gen.choose(0,x)
} yield (x,y)

def intPairInlined : (Gen.Parameters, Seed) =>
(Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
  }
\end{lstlisting}
\caption{Simplifying a generator in Scala}\label{fig:simplify-scala}
  \end{subfigure}
  \caption{Simplifying Generators}\label{fig:simplify}
\end{figure}

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking bug-finding speed}
\bcp{Make sure that capitalization of section titles is consistent}
To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
case studies using Etna~\cite{etna}.
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test a specific aspect of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations).
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.
Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
expect the results in this section to extend to \scalaname.

@@

In the years since the original QuickCheck paper~\cite{quickcheck00}, PBT has had remarkable success
in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages---including non-strict or procedural ones---might be curious about how the
techniques brought to bear here on OCaml's \bq{} and Scala's ScalaCheck
might be imported to their favorite language.

@@

  \textbf{Haskell} is home to the original PBT implementation, and
  so it is natural to ask why we have not applied \name{} to it yet. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) is negligible.  For
  complex enough programs, however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system, Template Haskell~\cite{templatehaskell}, that is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to build an \name{} version of Haskell
  QuickCheck, but the benefits would be much less pronounced.

  \textbf{F\#} has both multi-stage programming capabilities~\cite{fsmeta} and a well-used PBT library called FsCheck~\cite{fscheck}.
  The internals of FsCheck's generator eDSL are very similar to both
  \bq and ScalaCheck, so we expect that building \name{}-style generators
  in F\# should be a straightforward engineering effort.
  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations

  \textbf{Rust} shares some similarities with functional languages, and,
  indeed, it features a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However, Rust does not
  structurally encourage monadic eDSLs---it does not have a special monadic syntax---and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of
  the PBT libraries in Rust.

@@

Reordering\bcp{``reordering'' sounds like rearranging a deterministic
  order; this is changing the shape of a distribution, right?} test input sampling is perhaps the most well-studied way to accelerate bug-finding in PBT.
 Feedback-driven mechanisms
like Targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
Coverage-Guided PBT~\cite{lampropoulosCoverageGuidedProperty2019}
speed up testing by changing the generation order to find
``interesting'' inputs faster. Separately, enumerative approaches to
PBT~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
try to get to bugs faster by leveraging the ``small-scope
hypothesis,'' which posits that most bugs will be triggered by smallish
inputs. All of these approaches can have significant performance
benefits in
practice, but they are largely orthogonal to our contributions; we expect
\name{} could be used to speed these existing techniques to varying
degrees. \bcp{Namely: the staging part would be equally effective, but
the RNG part would be either less or not relevant.}

Similarly orthogonal to our approach are techniques for quickly filtering out
invalid inputs. Some of these approaches work statically,
by automatically deriving generators that produce only valid inputs by
construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
filter dynamically via laziness~\cite{claessenGeneratingConstrainedRandom2015}
or by solving satisfiability
problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
In all cases, these approaches could be further improved using
techniques from \name. \bcp{Why not also RNG improvements?  (I think
  \name{} refers {\em only} to the staging part of what we've done,
  but I'm not certain about that, and I'm not certain we've been consistent...)}

The most direct comparison to our work is
QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}, an
implementation of Haskell's QuickCheck that exploits the inherent parallelism of
PBT to achieve significant performance gains. QuickerCheck, like \name, was
designed with performance engineering in mind, taking seriously the idea that
PBT is a performance critical task. Still, despite their similar
motivations, the solutions are entirely different and complementary.

\paragraph{Multi-Stage Programming}
% \jwc{A generator is a parser of
% randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
% lots of staged parser libraries!}
Staging's roots come from quasiquotation in LISP \cite{bawden99}, where the
unified representation of data and code allows for sophisticated
metaprogramming. Quasiquotation-style macros were introduced to the ML family of
languages by Nielson and Nielson's Two-Level Functional Languages~\cite{nielson92} and
MetaML~\cite{taha00}, which in turn quickly inspired MetaOCaml~\cite{metaocaml} and more recently
MacoCaml~\cite{macocaml}, Lightweight Modular Staging in Scala~\cite{lms}, Template
Haskell in Haskell~\cite{templatehaskell}, and more.  In addition to implementations, researchers
have studied the type-theoretic foundations of multi-stage languages \cite{cmtt,
davies-pfenning, davies17}, as well as ways of combining multi-stage types with
other sophisticated features like dependent types \cite{kovacs24}. Meanwhile,
the Racket community continues the LISP tradition with a long and fruitful
line of work studying how macros and metaprogramming can be used to build extensible
DSLs \cite[etc.]{flatt12, samth11, skthesis}.

In ML-like languages, the primary use of staging is building embedded DSLs
with minimal performance overhead, an idea that has come to be known as ``abstraction without regret''
\cite{lms,parreaux17,parreauxthesis, sheard99, trattdsl}.
This technique has been applied across a range of domains, including
big-data processing \cite{jet12}, stream processing \cite{moller20,strymonas},
query processing \cite{rhyme}, and parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap}.
This work on staged parsers is the closest analogue to ours---indeed, PBT research has drawn an explicit link between parsing and generation, framing generators
as ``parsers of randomness'' \cite{goldsteinParsingRandomness2022}.
However, this equivalence is primarily theoretical. PBT generators (a)
are not actually implemented as parsers
of random sequences and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
For this reason, we must build staged generator libraries from scratch
and cannot simply use existing staged parser libraries.  \bcp{Maybe
  cut that last sentence?}

@@

\section{Conclusion \& Future Work}
\label{section:concl}
We have identified, studied, and proposed general solutions to two
important sources of inefficiency in PBT generator libraries: abstraction
overhead and choice of \rand{}.

In the future, we hope to continue to investigate and push the boundaries of
PBT generator performance.
%
On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion~\cite{kiselyov2022letrecinsertioneffects} that yields optimal let placement,
which we did not use in \camlname{}.
Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions\bcp{citation?}. Currently, the \camlname{}
recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this
overhead can be eliminated using a type-level heterogenous list to represent the state at compile time.

On the randomness side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cite{lehmerrng}, WyHash~\cite{wyrand},
and variants of xoroshiro~\cite{xorshiro}. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
or generating large buffers of random numbers ahead of time. Finally, we hope to investigate the degree to which statistical randomness
matters in PBT, since some of these ideas trade statistical guarantees for speed.
